---
title: "Exercise 3"
output: 
  html_document: 
    keep_md: yes
---


```{r}
#You get the temperatures recorded in 20 city weather stations of Germany on Dec 25 2013
temp = c(0,0,0,-2,-4,-5,-2,+1,-15,-3,-10,-10,+8,+7,+5,-20,0,-5,+6,+1)

# you also have the number of persons injured (sliding on an icy curb) the same day in a buffer area of 2 km around each weather station

injured = c(3,3,2,1,3,5,2,0,7,3,6,6,1,1,1,9,4,3,0,0)

```

(1)combine the 2 vectors into a dataframe called Germany; column names should be Temperatures and Injured

```{r}

```

(2) correlation or causation?
(3) if your answer to question (2) correlation, then test for correlation and report your results including a plot.
(4) if your answer to question (2) causation, then test for it  and report your results including a plot with your predictions (you cannot use effects library). You are required to include the quadratic term of your predictor and eventually select the best model using step AIC. 

```{r}


```

(5) do we meet model assumptions?

```{r}

```


