---
title: "Classical Tests Part III"
output: 
  html_document: 
    keep_md: yes
---

## Two dependend samples test. Wilcoxon signed rank test (NON PARAMETRIC)
This is a non-parametric method appropriate for examining the median difference in observations for 2 populations that are paired or dependent on one onother

Open the dataset 'streams'
Water measurements taken at different levels of a river:
'up' and 'down' are water quality measurements of the same river taken before and after a water treatment filter, respectively 

```{r}
setwd("~/TEACHING IN FREIBURG/11 - Statistics with R fall 2015/4_Classical Tests")
streams = read.table("streams.txt",header=T)
head(streams)
```



```{r}
attach(streams)
summary(streams)
plot(up, down)
abline(a = 0, b = 1) #add a line with intercept 0 and slope 1

```
The line you see in the plot corresponds to x = y, that means same water measuremets before and after the water treatment (it looks like to be true in 2 rivers only)


Our null hypothesis is that the median before and after is not different

```{r}
shapiro.test(down)
shapiro.test(up)
#the assumption of normality is certainly not met for measurements after the treatment. Let's go for a more conservative non-parametric procedure

summary(up)
summary(down)
boxplot(up, down)
wilcox.test(up, down, mu = 0, paired = T, conf.int = T, exact = F) #paired = T !!!
detach(streams)
```


## Two dependend samples test. Paired T-test (PARAMETRIC)
This is the parametric method for examining the difference in means for two populations that are paired or dependent on one another

```{r}
fish = read.table("fishing.txt",header=T)
```

This is a dataset about the density of a fish prey species (fish/km2) in 121 lakes before and after removing a non-native predator

  
```{r}
attach(fish)
head(fish)
boxplot(before,after,ylab="Fish Density",
        names=c("before", "after"))

shapiro.test(before)
shapiro.test(after)

plot(before,after)
abline(a=0,b=1) 
```


```{r}
t.test(before, after, mu = 0, paired = T)

t.test(after, before, mu = 0, paired = T) #changing the order of variables, we have a change in the sign of the t-test estimated mean of differences

detach(fish)
```

+++++++++++++++++
Edited by Simone Ciuti, University of Freiburg, 20/10/2015; 
Intended for the only purpose of teaching @ Freiburg University; 

+++++++++++++++++++++++++++++++++++++++++++++++++


