---
title: "GLM binomial"
output: 
  html_document: 
    keep_md: yes
---

Presence / absence data - binary data.

Logistic regression is a method for fitting a regression curve, y = f(x), when y consists of proportions or probabilities, or binary coded (0,1--failure,success) data. When the response is a binary (dichotomous) variable, and x is numerical, logistic regression fits a logistic curve to the relationship between x and y. 


Assumptions
The true conditional probabilities are a logistic function of the independent variables.
No important variables are omitted.
No extraneous variables are included.
The independent variables are measured without error.
The observations are independent.
The independent variables are not linear combinations of each other.

##Example 1 - tubercolosis in Wild Boar

Wild boar with/without tubercolosis as a function of body lenght (age is collinear with length and we do not use it as predictor)
```{r}
#import wildboar data
load("Boar.RData")
head(Boar)
summary(Boar)
cor(Boar[,c(3,4)])
```

A bit of data exploration
```{r}
attach(Boar)
boxplot(LengthCT ~ Tb, xlab = 'Body length', ylab = 'Tb', horizontal = T) #certainly, not the best way to show binomial data 

plot(Boar$LengthCT, Boar$Tb, xlab = "Body Length", ylab = "Tb")

#just to get a feeling with what is going on with these data
B0 = lm(Tb ~ LengthCT, data = Boar)
abline(B0) 
#wrong in principle, because we deal with probabilities here, while linear regression also predicts negative probabilities
```
We need a function that maps the values between 0 and 1
(e.g., logit link, probit link, clog-log link, log-log link).
the default is logit link

```{r}
B1 = glm(Tb ~ LengthCT, family = binomial, data = Boar) # logit link is the default
summary(B1)
```

Probability that an animal is infected is:
p = (e^(-3.89 + 0.03 * length)) / (1 + e^( -3.89 + 0.03 * length))

```{r}
plot(Boar$LengthCT, Boar$Tb,xlab="Length", ylab="Tb")
MyData = data.frame(LengthCT = seq(46.5, 165, by = 1))
Pred = predict(B1,newdata = MyData, type = "response", se = T)
lines(MyData$LengthCT, Pred$fit ,col = 2,lty = 2)
lines(MyData$LengthCT, Pred$fit + (1.96 * Pred$se.fit),col = 4,lty = 2)
lines(MyData$LengthCT, Pred$fit - (1.96 * Pred$se.fit),col = 4,lty = 2)


```



other option (actually changing the link function)
```{r}
B1.A = glm(Tb ~ LengthCT, family = binomial(link = "probit"), data = Boar)
summary(B1.A)
# logit and probit link assume that you have about the same number of zeros and ones
# clog-log is a good option if you have lots of zeros, or viceversa family = binomial(link="cloglog")

plot(Boar$LengthCT, Boar$Tb,xlab="Length", ylab="Tb")
MyData = data.frame(LengthCT = seq(46.5, 165, by = 1))
Pred = predict(B1.A,newdata = MyData, type = "response", se = T)
lines(MyData$LengthCT, Pred$fit ,col = 2,lty = 2)
lines(MyData$LengthCT, Pred$fit + (1.96 * Pred$se.fit),col = 4,lty = 2)
lines(MyData$LengthCT, Pred$fit - (1.96 * Pred$se.fit),col = 4,lty = 2)




```

Pseudo R Square
```{r}
#Null Deviance - residual Deviance / Null Deviance
((B1.A$null.deviance-B1.A$deviance)/B1.A$null.deviance)*100 #5% of total variability explained
```



##Example 2 - Parasites in cod (fish)


```{r}
ParasiteCod <- read.delim("ParasiteCod.txt")

head(ParasiteCod)
```
Response variable: Prevalance 0 1
Explanatories: year, area, depth at which fish were caught
Depth and area are collinear; other explanatory are sex, length, weight, stage, age..all collinear except for sex

```{r}
ParasiteCod$fArea = factor(ParasiteCod$Area)

attach(ParasiteCod)
P1 = glm(Prevalence ~ fArea + Length,
               family = binomial, data = ParasiteCod)  

summary(P1)

plot(Length,Prevalence)

MyData = data.frame(Length = seq(from = 20, to = 100, by = 1), fArea="1")
Pred = predict(P1,newdata = MyData, type = "response", se = T)
lines(MyData$Length, Pred$fit, col = 1,lty = 1)
lines(MyData$Length, Pred$fit + (1.96 * Pred$se.fit), col = 1,lty = 3)
lines(MyData$Length, Pred$fit - (1.96 * Pred$se.fit), col = 1,lty = 3)


MyData = data.frame(Length = seq(from = 20, to = 100, by = 1), fArea="2")
Pred = predict(P1,newdata = MyData, type = "response", se = T)
lines(MyData$Length, Pred$fit, col = 2,lty = 1)
lines(MyData$Length, Pred$fit + (1.96 * Pred$se.fit), col = 2,lty = 3)
lines(MyData$Length, Pred$fit - (1.96 * Pred$se.fit), col = 2,lty = 3)




MyData = data.frame(Length = seq(from = 20, to = 100, by = 1), fArea="3")
Pred = predict(P1,newdata = MyData, type = "response", se = T)
lines(MyData$Length, Pred$fit, col = 3,lty = 1)
lines(MyData$Length, Pred$fit + (1.96 * Pred$se.fit), col = 3,lty = 3)
lines(MyData$Length, Pred$fit - (1.96 * Pred$se.fit), col = 3,lty = 3)



MyData = data.frame(Length = seq(from = 20, to = 100, by = 1), fArea="4")
Pred = predict(P1,newdata = MyData, type = "response", se = T)
lines(MyData$Length, Pred$fit, col = 4,lty = 1)
lines(MyData$Length, Pred$fit + (1.96 * Pred$se.fit), col = 4,lty = 3)
lines(MyData$Length, Pred$fit - (1.96 * Pred$se.fit), col = 4,lty = 3)



# you can fix it with a loop, if you wish
AREA = 1:4
plot(Length,Prevalence)
for (i in AREA) {
  MyData = data.frame(Length = seq(from = 20, to = 100, by = 1), fArea = as.factor(i))
Pred = predict(P1,newdata = MyData, type = "response", se = T)
lines(MyData$Length, Pred$fit, col = i,lty = 1)
lines(MyData$Length, Pred$fit + (1.96 * Pred$se.fit), col = i,lty = 3)
lines(MyData$Length, Pred$fit - (1.96 * Pred$se.fit), col = i,lty = 3)
}

legend("bottomright", c("Area 1", "Area 2", "Area 3", "Area 4"), col = 1:4, title = "",
       bty="n", pch = rep("_", 4), inset = .04, cex = 1.2)



```




##Example 3 - Tubercolosis in deer, this time proportion data (ranging from 0 to 1)

```{r}
Tbdeer <- read.delim("Tbdeer.txt")
head(Tbdeer)
Tbdeer$DeerPosProp = Tbdeer$DeerPosCervi / Tbdeer$DeerSampledCervi  # proportion of deer infected

Tbdeer$fFenced = factor(Tbdeer$Fenced)
```


Effect of percentage open land, scrubs, pine plantation, number of quercus per unit area, red deer abundance index, estate size, estate fenced ON probability of getting infected

```{r}
Deer2 <- glm(DeerPosProp ~ OpenLand + ScrubLand + ReedDeerIndex + fFenced,
        family = binomial, weights = DeerSampledCervi, data = Tbdeer)

summary(Deer2)


```


This is a Binomial GLM with proportion data.
We can have overdispersion. 
Residual deviance / df should be ~1

```{r}
Deer2$deviance / Deer2$df.residual #~9

library(effects)
plot(allEffects(Deer2))
```

we add a dispersion parameter to the variance of Y
```{r}
Deer3 <- glm(DeerPosProp ~ OpenLand + ScrubLand + ReedDeerIndex + fFenced,
             family = quasibinomial, weights = DeerSampledCervi, data = Tbdeer)
summary(Deer3)
Deer3$deviance / Deer3$df.residual

plot(allEffects(Deer3))
```

We have lost many of the significant patterns that were biased by overdispersion.
The new results are pretty different.


```{r}

plot(Tbdeer$OpenLand, Tbdeer$DeerPosProp)

MyData = data.frame(OpenLand = seq (min(Tbdeer$OpenLand), max(Tbdeer$OpenLand), 0.05),
                    ScrubLand = mean(Tbdeer$ScrubLand), ReedDeerIndex = mean(Tbdeer$ReedDeerIndex), fFenced = "0") # scenario no-fenced
Pred = predict(Deer3,newdata = MyData, type = "response", se = T)
lines(MyData$OpenLand, Pred$fit, col = 3,lty = 1)
lines(MyData$OpenLand, Pred$fit + Pred$se.fit, col = 3,lty = 3)
lines(MyData$OpenLand, Pred$fit - Pred$se.fit, col = 3,lty = 3)



MyData = data.frame(OpenLand = seq (min(Tbdeer$OpenLand), max(Tbdeer$OpenLand), 0.05),
                    ScrubLand = mean(Tbdeer$ScrubLand), ReedDeerIndex = mean(Tbdeer$ReedDeerIndex), fFenced = "1") # scenario yes-fenced
Pred = predict(Deer3,newdata = MyData, type = "response", se = T)
lines(MyData$OpenLand, Pred$fit, col = 1,lty = 1)
lines(MyData$OpenLand, Pred$fit + Pred$se.fit, col = 1,lty = 3)
lines(MyData$OpenLand, Pred$fit - Pred$se.fit, col = 1,lty = 3)



```

-> IN-class discussion about diagnostics in GLM binomial and model evaluation.

+++++++++++++++++
Edited by Simone Ciuti, University of Freiburg, 27/10/2015; 
Intended for the only purpose of teaching @ Freiburg University; 
Sources: Mick Crawley, R book, 2nd edition; Simone Ciuti, simulated data.; Mike Marin Stat, University of British Columbia; quick-R; 
Zuur, Ieno, Walker, Saveliev and Smith 2009 Mixed Effects Models and Extensions in Ecology with R.
+++++++++++++++++++++++++++++++++++++++++++++++++

