---
title: "1 - Setup R & RStudio"
output: 
  html_document: 
    keep_md: yes
---

R & R Studio are open source. 
http://www.r-project.org/
http://www.rstudio.com/

## Basic operations, objects, vectors and matrices.

Open R Studio which runs R

Let's create and save an R studio (very simple) script (on the RStudio command bar: File -> New File -> R script) 
Remember! You can use the symbol # to add comments that will not be processed by R as codes. 
Use of "Run" to run your script in the R console (otherwise, shortcut: Ctrl + enter)

Let's start with assigning values to objects in R, doing basic arithmetic functions, and performing a few other handy tasks in R.


```{r}
# assign a value to an object
x = 55
print(x)
#or
x
#remember R is case sensitive - try to type X
#Error: object 'X' not found

# R does not care about SPACING, although it is good habit to put some space in your codes because they will be more readable eventually
a = 1
b=1

# use of arrow instead of "=". Remember, the arrow is directional (you can use both ways), while "=" assigns a value (on the rigth of =) to the object name (on the left of =)

z = 55 # assing 55 to z
# 55 = z does not work - try it out in your console

y <- 14
23 -> zz

```

Check the top righ window to verify what we just stored in the workspace memory.


or use the ls command
```{r}
ls()
```

you can remove an object
```{r}
rm(y)
```


Object names can include numbers
```{r}
z1 = 15
z1
#but cannot begin with numbers
# try to type 2z =15

```

Quotaion marks to assign character to objects
```{r}
m1 = "Rclass"
m1

m2="25" #this is a character, not a number!
m2
```

```{r}
2 + 2

#but also with our objects
z1+x

#we store it in a new object
z = z1 + x
z


#we can square z
z^2

sqrt(z) + x^2


#natural log
log(x)

#log base 2
log2(x)

f = -34
abs(f)


```

common mistake in R (incomplete command)
```{r}
# try to type sqrt(x
# you gotta complete the script!  (or press ESC if you are stuck)
```

- few shorcuts
In the R consol, try the arrow up key command


let's create a vector using the concatenate command
```{r}
x1 = c(1,3,5,7,9)
x1

gender = c("male", "female")
gender
#sequence
2:7

seq(from = 1, to = 7, by = 1)


seq(from = 1, to = 7, by = 1/3)


seq(from = 1, to = 7, by = 0.25)
```

Let's create a vector repeating something for a certain number of times

```{r}
rep(1, times = 10)

rep("simone", times = 10)

rep(1:5, times=5)

rep(seq(from = 2, to = 5,by = 0.25), times = 5)

rep(c("m","f"),times=10)
```



```{r}
x = 1:5
y = c(1, 3, 5, 7, 9)

```

we can add a value to each element of the vector
```{r}
x + 10

x - 10


x * 10
```

if 2 vectors are of the same length
we may add/subtract/mult/div corresponding elements
```{r}
x
y
x + y
x - y
```

We can extract elements of a vector using squared brakets
```{r}
y
y[2]

#negative sign extract all except that element
y[-2]

y[1:3]

#extract the first anf the third
y[c(1,3)]

#all except them
y[-c(1,3)]

# you can include a conditional query
y[y < 3]
y[y < 20]

```

we can create a matrix of value using the matrix command

```{r}
matrix(c(1,2,3,4,5,6,7,8,9), nrow = 3, byrow = TRUE) #enter the elements rowwise

matrix(c(1,2,3,4,5,6,7,8,9), nrow = 3, byrow = FALSE)#values entered columnwise



mat1 = matrix(c(1,2,3,4,5,6,7,8,9), nrow=3, byrow=TRUE) #enter the elements

mat1
```

square brakets used to grab elements from the matrix

```{r}
mat1[1,2]  #element in the first row and second column

mat1[c(1,3),2]

mat1[2,] #row 2 and all columns

mat1[,1]


mat1*10


# convert the matrix into a dataframe
mat1 = data.frame(mat1)
names(mat1) = c("column1", "column2", "column3")
mat1
mat1[,3] # selecting data on column3, all rows


# shortcut R studion Ctrl + L to clean up your console (only visualization, object are not deleted!)
```

-> browse the HTLM versione produced by Knitting this MarkDown file.
-> IN CLASS EXERCISES 1.EXERCISES.txt

+++++++++++++++++
Edited by Simone Ciuti, University of Freiburg, 14/10/2015; 
Intended for the only purpose of teaching @ Freiburg University.
+++++++++++++++++++++++++++++++++++++++++++++++++