---
title: "Few more R Studio tips"
output: 
  html_document: 
    keep_md: yes
---

Working directory: how to save your working directory [usually not recommended -> it is always good habit to re-run the scripts at the beginning of each session. If certain operations require long processing time, then always better to save the files / models and load them, rather than saving the working directory. (very personal choice though)
each project = 1 working directory



```{r}
mtcars -> mtcars #this is just to add the object to our workspace


getwd() # to know our current working directory
setwd("~/TEACHING IN FREIBURG/11 - Statistics with R fall 2015/1_intro to R and R Studio") # to set my current wd

```



```{r}
meanhp = mean(mtcars$hp)
y=c(1,2,3)
z=15
objectsum = summary(mtcars)
objectsum


```


OK, say you want to quit here. You may want to save the workspace to be able to continue exactly from where you left.
```{r}
#option 1
save.image("myfirstproject.Rdata")

#option 2 Session menu -> save workspace as

#option 3
#when you quit R studio, you'll be asked whether you wish to save the workspace

```

```{r}
#clean the workspace
rm(list=ls())    #or, menu -> session -> clear workspace

#quit R studio
```

Reopen R studio
```{r}
# set the directory with the script or the scroll menu
setwd("~/TEACHING IN FREIBURG/11 - Statistics with R fall 2015/1_intro to R and R Studio") 
load("myfirstproject.Rdata")

#alternative menu -> Session -> load workspace
```




tab key... you start to type an R command (e.g. summary)
also to recall an object name (e.g. obj)


##install packages
packages are add-ons that can extend R's functionality and perform specific tasks covering a wide range of modern statistic

```{r}
#install.packages("epiR")

#but see the menu bar -> Tools -> install packages
library(epiR)
library(effects)
```

-> END OF TOPIC.

+++++++++++++++++
Edited by Simone Ciuti, University of Freiburg, 9/10/2014; 
Intended for the only purpose of teaching @ Freiburg University
+++++++++++++++++++++++++++++++++++++++++++++++++
