---
title: "Intro to exploratory stat (basic plotting)"
output: 
  html_document: 
    keep_md: yes
---

```{r}
setwd("~/TEACHING IN FREIBURG/11 - Statistics with R fall 2015/2_intro to very basic exploratory statistics") #set working directory
```


## Tips to better handle plots in R
graphics.off()  # shuts down all open graphics devices.
x11(h = 12, w = 12)  # opens a new plotting window 12 by 12
plot(mtcars)
savePlot(filename="myplot",type="wmf", device=dev.cur()) # save the plot (metafile), but it can be a pdf, bmp ect.
savePlot(filename="myplot1",type="pdf", device=dev.cur())




## Bar Charts 
appropriate for summarizing the distribution of a categorical variable

```{r}
# basic data handling - let's convert am into a factor (new column fam)
head(mtcars)
mtcars$fam=factor(mtcars$am, levels=c(0,1), labels=c("automatic","manual"))
head(mtcars)


```


```{r}
attach(mtcars)
table(fam) #frequency table

#let's store it in a object
count = table(fam)
count
#let's compute frequencies %
percent = table(fam) / length(fam)
percent

barplot(count)

barplot(percent)


#let's add a title
barplot(percent, main="the title", xlab="transmission", ylab="%", las=1, ylim=c(0,1), names.arg=c("aut transm", "man transm") )

```

## Pie Charts

```{r}
pie(count, main="the title", col=c(2,3))
box()
```

?pie 
?barchart

-> homework: change the look of these plots. just google it!

## boxplot
appropirate to summarize the distribution of a numerical variable
type ?boxplot

```{r}
summary(mpg)
quantile(mpg)
quantile(mpg,probs=c(0, 0.25, 0.50, 0.75,1))

boxplot(mpg,main="title", ylab="mpg", ylim=c(0,50), las=1)

boxplot(mpg~fam, main="MPG by transmission")

boxplot(mpg[fam=="automatic"])
```



## histograms
appropriate for summarizing the distribution of a numerical variable 
WARNING(!!) -> depending on the number of breaks (or bins), histograms look very differently. Not always a good tool! 

```{r}
hist(mpg)
hist(mpg, freq=FALSE, main="my histogram")
hist(mpg, prob=TRUE, ylim=c(0,0.1))
lines(density(mpg))  #add a density curve
lines(density(mpg), col="red", lwd=3)
hist(mpg, breaks = seq(10,36, by=2), las=1)

```


## scatterplots
appropriate to examine the relation between two numerical variables
 
 
Relationship between horsepower hp and consumption mpg
```{r}
plot(mpg ~ hp) # y~x form
plot(hp, mpg) # x,y form

plot(hp, mpg, xlab = "Gross horsepower", ylab="Miles/(US) gallon", las = 1, col = "red", xlim = c(0,400), cex = 2 )  #cex (plotting characters size times 2)

plot(hp, mpg, xlab = "Gross horsepower", ylab="Miles/(US) gallon", las = 1,col = "red", xlim = c(0,400), pch = 8)  #pch (plotting symbol)

abline(lm(mpg ~ hp), col="blue", lwd=5, lty=3) #add linear regression, line width and type



x = 1:20
plot(x, pch = x, cex = 3)
plot(x, col = x, cex = 3)
plot(x, col = x, cex = x)

```


## summary statistics

categorical variable fam
```{r}
table(fam)

table(fam)/length(fam)

#contingency table (2-way table)
table(mpg,fam)

table(fam,mpg)




```

numerical variable mpg
```{r}
mean(mpg)
summary(mpg)
sd(mpg)
var(mpg)
sqrt(var(mpg)) # = to sd
sd(mpg)^2 # = to variance
max(mpg)

tapply(mpg,fam,mean)
tapply(mpg,list(fam,gear),mean)
```


-> IN CLASS EXERCISES 1_introEXERCISES.txt

+++++++++++++++++
Edited by Simone Ciuti, University of Freiburg, 9/10/2014; 
Intended for the only purpose of teaching @ Freiburg University
+++++++++++++++++++++++++++++++++++++++++++++++++