---
output:
  html_document:
    keep_md: yes
    fig_width: 5
---
---
layout: page
title: Explorative data analysis
category: stats
subcategory: Descriptive statistics
---

```{r, echo=F}
library(knitr)
opts_knit$set(global.par=TRUE) 
opts_chunk$set(cache.extra = rand_seed,fig.align='center')
set.seed(2)
```


Explorative data analysis has the goal to explore and describe the patterns in your data, without any particular hypothesis. 


## Good introductions on this topic

* See further discussion in the [Essential Statistics](https://www.dropbox.com/s/8fmh10fdn6jd2xb/EssentialStatistics.pdf?dl=0) lecture notes, chapter on descriptive statistics as well as appendix on graphics and summary statistics

* For more details about plotting, please visit: [http://biometry.github.io/APES/R/R40-plottingInR.html](http://biometry.github.io/APES/R/R40-plottingInR.html)


## Some examples in R

### Exploring categorical data 


#### Summary statistics

Some piece of information that gives a quick and simple description of the data.

```{r}
attach(mtcars)
head(mtcars)
fam=mtcars$fam=factor(mtcars$am, levels=c(0,1), labels=c("automatic","manual"))
#we substract the variable Transmission (0 = automatic, 1 = manual) with "$" from the dataset 
head(mtcars)
```

Frequency table of the Transmission variable
```{r}
table(fam)
count=table(fam) 
count
```

% frequencies calculation
```{r}
percent=table(fam)/length(fam)
percent
```


#### Bar charts

Bar charts are appropiate to summarize categorical variables distributions

```{r}
barplot(percent, main="Percentage of cars with / without transmission", xlab="transmission", ylab="%", las=1, ylim=c(0,1), names.arg=c("auto transm", "manual transm") )
```


### Exploring continous / categorical data



#### Summary statistics


For a numerical variable, like "mpg"
```{r}
mean(mpg)
summary(mpg)
sd(mpg) #standard deviation
var(mpg) #variance
sqrt(var(mpg)) # = to sd
sd(mpg)^2 # = to variance
max(mpg)

tapply(mpg,fam,mean)
tapply(mpg,list(fam,gear),mean)
```

#### Boxplot 

Boxplots are appropiate to summarize numerical variables distributions
```{r}
summary(mpg)
quantile(mpg)
quantile(mpg,probs=c(0,0.20,0.40,0.60,0.80,1))

boxplot(mpg~fam, main="mpg by transmission")
```

#### Histograms 

Histograms are appropiate to summarize numerical variables distributions
```{r}
hist(mpg)
```

#### Stem and Leaf Plots 

Stem and Leaf plots are appropiate to summarize numerical variables distributions (low sample size)
```{r}
stem(mpg)   
```
`?stem` for more info
There are 2 obs 10.4
There is one obs 32.4 and one 32.9
 
#### Scatterplots 

Scatterplots are appropiate to summarize the relation between two numerical variables 

Relation ship between horsepower hp and consumption mpg
```{r}
plot(mpg~hp) # y~x
plot(hp, mpg) # x,y
plot(hp, mpg,xlab = "Gross horsepower", ylab="Miles/(US) gallon", las=1, col="red", xlim=c(0,400), cex =2 )
#cex (plotting characters size times 2)
```




