---
output:
  html_document:
    keep_md: yes
---
Generalized linear models (GLM)
===


# The general idea

General ideas of linear regression is that 

* Response is continous, theoretically from -infinity to + infinity
* Residuals are normally distributed around the model predictions

Idea of the GLM framework is take the linear regression framework, but allow relaxing both assumptions. To do this, we have to do two things

* We wrap the linear model in a transformation function that forces the response on the right interval (typical intervals are positive, or between 0 and 1). This transformation is called the link function
* We use other distributions as the Gaussian for the fit.

Those two ideas are explained in more detail below


## Other distributions

## Link function


# Important GLM types

## Logistic Regression


```{r}

## binomial data
require(effects) # holds the data
head(TitanicSurvival)
attach(TitanicSurvival)
 
plot(survived ~ age)
surv <- as.numeric(survived)-1
 
plot(surv ~ age)
 
fmt <- glm(surv ~ age + I(age^2) + I(age^3), family=binomial)
summary(fmt)
 
 
# showing that residuals are not normal at all
#hist(fmt$residuals, breaks = 100)
#abline(v = 0, col = "red", lwd = 6, lty = 2)
# residual plots can also be created with plot(fmt)
 
 
newage <- seq(min(age, na.rm=T), max(age, na.rm=T), len=100)
preds <- predict(fmt, newdata=data.frame("age"=newage), se.fit=T)
lines(newage, plogis(preds$fit), col="purple", lwd=3)
lines(newage, plogis(preds$fit-2*preds$se.fit), col="purple", lwd=3, lty=2)
lines(newage, plogis(preds$fit+2*preds$se.fit), col="purple", lwd=3, lty=2)
 
 
fmt <- glm(surv ~ age + sex + passengerClass, family=binomial)
summary(fmt)
 
detach(TitanicSurvival)

```



## Poisson Regression


```{r}
## Poisson:
 
# data
cfc <- data.frame(
  stuecke = c(3,6,8,4,2,7,6,8,10,3,5,7,6,7,5,6,7,11,8,11,13,11,7,7,6),
  attrakt = c(1,1,1,1,1,2,2,2,2,2,3,3,3,3,3,4,4,4,4,4,5,5,5,5,5) 
)
attach(cfc)
plot(stuecke ~ attrakt)
 
fm <- glm(stuecke ~ attrakt, family=poisson)
summary(fm)
 
newattrakt <- c(1,1.5,2,2.5,3,3.5,4,4.5,5)
preds <- predict(fm, newdata=data.frame("attrakt"=newattrakt))
lines(newattrakt, exp(preds), lwd=2, col="green")
 
# same with 95% confidence interval:
preds <- predict(fm, newdata=data.frame("attrakt"=newattrakt), se.fit=T)
str(preds)
lines(newattrakt, exp(preds$fit), lwd=2, col="green")
lines(newattrakt, exp(preds$fit+2*preds$se.fit), lwd=2, col="green", lty=2)
lines(newattrakt, exp(preds$fit-2*preds$se.fit), lwd=2, col="green", lty=2)
 
detach(cfc)

```


# Technical details 


```{r}
## Poisson with optimization by hand
 
# data
cfc <- data.frame(
  stuecke = c(3,6,8,4,2,7,6,8,10,3,5,7,6,7,5,6,7,11,8,11,13,11,7,7,6),
  attrakt = c(1,1,1,1,1,2,2,2,2,2,3,3,3,3,3,4,4,4,4,4,5,5,5,5,5) 
)
attach(cfc)
plot(stuecke ~ attrakt)
 
# defining the likelihood of a Poisson regression
 
loglikelihood <- function(par, independent = attrakt, observed = stuecke){
  linear = par[1]*independent + par[2]  # linear predictor
  predict = exp(linear)                 # link function
  logprobabilities = dpois(observed, predict, log=T)  # distribution
  return(-sum(logprobabilities))
}
 
# plotting the likelihood for different slopes, fixed intercept
 
slope=seq(0.05,0.25,length.out = 100)
intercept = rep(1.4,100)
pars = cbind(slope, intercept)
plot(slope, apply(pars,1, loglikelihood), main = "Likelihood profile", ylab = "Neg Log Likelihood")
 
# plotting the likelihood surface (= likelihood as a function of slope, intercept)
 
intercept = seq(1.2,1.7,length.out = 100)
parametervalues = expand.grid(slope,intercept)
parametervalues$response = apply(parametervalues,1, loglikelihood)
contour(slope, intercept, matrix(parametervalues$response, nrow = 100), nlevels = 20, main = "Likelihood response surface", xlab = "slope parameter", ylab = "intercept parameter")
 
 
# optimization with the optim function
 
bestfit = optim(c(0.12,1.3), loglikelihood, method = "BFGS")
points(bestfit$par[1], bestfit$par[2], col = "red", lwd = 4)
 
# calculating the "Hessian" matrix (Funktionaldeterminante)
 
bestfit = optim(c(0.12,1.3), loglikelihood, method = "BFGS", hessian = T)
bestfit$hessian
 
persp(slope, intercept, matrix(parametervalues$response, nrow = 100), theta = 40, phi = 30, expand = 0.7, col = c("grey","red"),  ticktype = "detailed")
 
# bonus plot
library(rgl)
persp3d(slope, intercept, matrix(parametervalues$response, nrow = 100), col = c("grey","red"), box = FALSE)
 
 
detach(cfc)

```





## Links

http://blog.revolutionanalytics.com/2014/04/some-r-resources-for-glms.html

