---
title: "Multiple linear regression and ANCOVA"
output: 
  html_document: 
    keep_md: yes
---

The multiple regression is an extension of the simple regression, where we simply add more numeric predictors.

```{r}
setwd("~/TEACHING IN FREIBURG/11 - Statistics with R fall 2015/8_ANCOVA")
load("lung.RData")

head(lung)
attach(lung)
```


```{r}
model1 = lm(LungCap ~ Age + Height)
summary(model1)
```

let's quickly check the predictions of the model using the library Effects
```{r}
library(effects)
plot(allEffects(model1))
plot(effect("Age",model1))
plot(effect("Height",model1))


```

let's now try the ANCOVA (we blend regression with ANOVA - in simple words, we have at least 1 numeric predictor and 1 categorical predictor)
```{r}
model2 = lm(LungCap ~ Age + Height + Smoke + Gender + Caesarean)
summary(model2)

plot(allEffects(model2))
```

```{r}
### check the assumptions
par(mfrow = c(2, 2))
plot(model2) #all good here
par(mfrow = c(1, 1)) 
detach(lung)
```

## Plotting predictions in ANCOVA

```{r}
# we manually plot the predictions of the following model:
model3 = lm(LungCap ~ Age + Smoke, data = lung)

# scenario 1: effect of Age on Lung Capacity for non-smokers
newdata = data.frame(Age = seq(min(lung$Age), max(lung$Age), 1), Smoke = "no")
head(newdata)
mypredictions = predict(model3, newdata = newdata, se = T)
plot(lung$Age[lung$Smoke == "no"], lung$LungCap[lung$Smoke == "no"], xlab = "Age", ylab = "Lung capacity", col = "grey", pch = 20)
lines(newdata$Age, mypredictions$fit, col = "grey", lwd = 3)
lines(newdata$Age, mypredictions$fit + (1.96 * mypredictions$se.fit), col = "gray")
lines(newdata$Age, mypredictions$fit - (1.96 * mypredictions$se.fit), col = "gray")



#scenario 2: effect of Age on Lung Capacity for smokers this time
with(lung, table(Age, Smoke)) # smokers occur when age >= 10. No reason to predict where we have no data (and babies are not smokers anyways)

newdata = data.frame(Age = seq(10, max(lung$Age), 1), Smoke = "yes")
head(newdata)
mypredictions = predict(model3, newdata = newdata, se = T)
points(lung$Age[lung$Smoke == "yes"], lung$LungCap[lung$Smoke == "yes"], col = "black", pch = 20, cex = 1.5)
lines(newdata$Age, mypredictions$fit, col = "black", lwd = 3)
lines(newdata$Age, mypredictions$fit + (1.96 * mypredictions$se.fit), col = "black")
lines(newdata$Age, mypredictions$fit - (1.96 * mypredictions$se.fit), col = "black")

legend("topleft", c("non-smokers", "smokers"), col = c("grey","black"), title="Smoking habit",
       bty="n", pch=c(20, 20),inset=.02, cex = 1.2)


# clearly. you can manually predict more complex ANCOVA models. If you had more predictors in your model, you would have to include them in the newdata as well, keeping them constant to mean (if numeric) or to a certain category (if categorical).

```


## Collinearity 
The collinearity is defined as a strong (over the limit) correlation between predictors, and it may lead to parameter instability. 
Are we allowed to fit Age + Height + Smoke + Gender + Caesarean as predictors in the same model?
For instance, we may expect that Heigth and Age are strongly correlated

```{r}
attach(lung)
plot(Height,Age)
cor(Height,Age) # certainly, Height and Age ar collinear and shouldn't be included as predictors in the same model.

# we cannot correctly say that the slope for age is the actual effect of age on 
# lung capacity adjusting for height age and height are bounded together! 

# our model structure was
#model2 = lm(LungCap ~ Age + Height + Smoke + Gender + Caesarean)
# we cannot include Height and Age in the same model due to collinearity issues!! (|r|>0.7)

model1 = lm(LungCap ~ Height); summary(model1) #adj r2 = 0.83
cor(Height, LungCap)

model2=lm(LungCap ~ Age); summary(model2) #adj r2 = 0.67
cor(Age,LungCap)



```


## Checking for and avoiding collinearity issues!

 
 
```{r}
names(lung)
detach(lung) #fist of all, we convert factor into numeric to check whether they are collinear with other variables too
lung$NSmoke = as.numeric(lung$Smoke)
lung$NGender = as.numeric(lung$Gender)
lung$NCaesarean = as.numeric(lung$Caesarean)
summary(lung)
```


2 different tools to check for collineariy
1) PAIRWISE SCATTERPLOTS with correlation coefficients
2) VIF VARIANCE INFLATION FACTORS (MULTICOLLINEARITY)

 
 
```{r}

source("collinearity check.r")  #loading useful functions

#bind together the column of interest
attach(lung)
Z = cbind(LungCap, Age, Height, NSmoke, NGender, NCaesarean)

pairs(Z, lower.panel = panel.smooth2,
      upper.panel = panel.cor, diag.panel = panel.hist)

# rule of thumb |r|>0.7 heavily collinear (however, 0.5 can be used as well -> see Dormann's book)

```

VIF  rule of thumb: VIF>3 highly collinear
```{r}
corvif(Z[,-1])  # we have to exclude the response variable from the VIF test
# again, problem here with Height and Age
# we decide to keep the Heigth and discard Age (based on the higher correlation with the response variable)
corvif(Z[,c(-1,-2)])

# ok, we are good. we can build our multiple regression with the other predictors
# response = "LungCap"
# indipendent variables = "Height"     "Smoke"      "Gender"     "Caesarean" 

# depending on your research question, you may want to include Age instead of height. The alternative is to create a new variable using PCA. See below.

```

## PCA (principal component analysis -> a good option to deal with collinear predictors)

```{r}
detach(lung)
newdata = lung[,c("Height", "Age")]
myPCA = prcomp(newdata, scale = TRUE)
summary(myPCA)     # the first axis PC1 explains 91% of variability of Height and Age


biplot(myPCA, main = "example of PCA ", cex = 0.5)
#   Age and Height have a positive loading on PC1 (first axis of the principla component)
#   Height has a positive loading on PCA2, Age has a negative loading on it. 

# finally, you attach PC1 and PC2 to your data

lung$PC1 = predict(myPCA)[,1]      
lung$PC2 = predict(myPCA)[,2]  
head(lung)

#eventually, you can use principal component axis loadings instead of Age or Height.
# this is very useful WHEN YOU HAVE MANY COLLINEAR PREDICTORS (WEATHER DATA, ENVIRONMENTAL PREDICTORS ECC) AND YOU DO NOT WANT TO PICK ONE. 

```





+++++++++++++++++
Edited by Simone Ciuti, University of Freiburg, 27/10/2015; 
Intended for the only purpose of teaching @ Freiburg University; 
Sources: Mick Crawley, R book, 2nd edition; Simone Ciuti, simulated data.; Mike Marin Stat, University of British Columbia; quick-R; Zuur et al Mixed Effects Models in R; Zuur et al Analysing Ecological Data 
+++++++++++++++++++++++++++++++++++++++++++++++++

