---
output:
  html_document:
    keep_md: yes
---
---
layout: page
title: Data in R
category: r
---

Before we can work with data, we need to be able to read it in an represent it in R

## Reading in data

* See tutorial on reading in data [here](http://www.statmethods.net/input/importingdata.html)

## Data structures

Important data structures in R are

* Atomic variables (numeric, char)
* Vectors (a sequence of atomic variables. For example c(1,2,3,4) is a vector of 4 numeric variables)
* Lists (like a vector, but unordered)
* A data frame (technically a list of vectors, untechnically think of it as a spreadsheat with collumns that can be of different atomic type)

Hint: you can find out the type and structure of an object in R with the `class()` and `str()` command

Tutorial and examples of how to use them are 

* Appendix 2 of the [Essential statistics](https://www.dropbox.com/s/8fmh10fdn6jd2xb/EssentialStatistics.pdf?dl=0) lecture notes

* This [tutoria](http://ecology.msu.montana.edu/labdsv/R/labs/R_ecology.html#recol)

* The [QuickR](http://www.statmethods.net/) reference

* The [R basics tutorial](http://www.uni-kiel.de/psychologie/rexrepos/rerR_Basics.html)

The following subsections only give an introduction


## Some basic examples about data structures in R

### Basic variable assignments

It is possible to assign a value to an object with the `=` or `->` symbols

```{r}
x = 55
x <- 55
```

and check the result with the `print` command

```{r}
print(x)
#or
x
```

R is case sensitive, if we try to type `X` we got the following message: `Error: object 'X' not found`

A common mistake in R is to write an incomplete command. If we type `sqrt(x` we get an error. We gotta complete the script (or press ESC if the programm gets stuck)


It is possible to remove an object from our workspace with the `rm` command 
```{r}
rm(y)
```

Object names can include numbers
```{r}
z1 = 15
z1
```

But they cannot begin with numbers. `2z =15`, for example, would create an error

We use quotation marks to assign characters to objects
```{r}
m1 = "Rclass"
m1

m2="25" #like this, 25 is not a number anymore, but a character
m2
```


### Vectors

We can create a vector by using the concatenated command `c()`
```{r}
x1 = c(1,3,5,7,9)
x1

gender = c("male", "female")
gender

2:7 # sequence from 2 to 7

seq(from=1, to=7, by=1) # sequence from 1 to 7 by 1

seq(from=1, to=7, by=1/3) # sequence from 1 to 7 by 0,333

seq(from=1, to=7, by=0.25) # sequence from 1 to 7 by 0,25
```

Create a vector repeating something a certain number of times
```{r}
rep(1, times=10)

rep("vector", times=10)

rep(1:5, times=5)

rep(seq(from=2, to=5,by=0.25), times =5)

rep(c("m","f"),times=10)
```

we can add a value to each element of the vector
```{r}
x = 1:5

x + 10
x-10
x*10
```

We may add/subtract/mult/div but the vectors HAVE to be the same length
```{r}
y = c(1,3,5,7,9)
x = 1:5

x
y
x+y
x-y
```

It is possible to extract elements of a vector by using squared brakets
```{r}
y
y[2]

#A negative sign indicates R to extract all the elements except that one
y[-2]

#extract the first and the third elements
y[c(1,3)]

#extract all the elemets except the first and the third
y[-c(1,3)]

#extract all the elements above the third one
y[y<3]

```

### Matrices

We can create a matrix of values by using the matrix command
```{r}
matrix(c(1,2,3,4,5,6,7,8,9),nrow=3,byrow=TRUE) #enter the elements rowwise

matrix(c(1,2,3,4,5,6,7,8,9),nrow=3,byrow=FALSE) #values entered columnwise

mat1= matrix(c(1,2,3,4,5,6,7,8,9),nrow=3,byrow=TRUE) #enter the elements

mat1
```

Square brakets are used to grab elements from the matrix
```{r}
mat1[1,2]  #element in the first row and second column

mat1[c(1,3),2]

mat1[2,] #row 2nd and all the columns

mat1[,1]

mat1*10
```

### Data frames

The data frame is the most common choice to represent your field data, thefore it's important to know how to work with them and select data from them. 

[http://www.uni-kiel.de/psychologie/rexrepos/rerData_Frames.html](http://www.uni-kiel.de/psychologie/rexrepos/rerData_Frames.html)


## Working with big data 

[http://blog.revolutionanalytics.com/2014/08/the-iris-data-set-for-big-data.html](http://blog.revolutionanalytics.com/2014/08/the-iris-data-set-for-big-data.html)

[Text File Splitter](http://www.systemwidgets.com/Downloads/FreewareDownloads/TextFileSplitter/tabid/120/Default.aspx): can split very larve (> several GB) textfiles.

# Coupling to databases 

* [http://www.win-vector.com/blog/2016/02/using-postgresql-in-r/](postgresql)

