---
title: "Normal distribution"
output: 
  html_document: 
    keep_md: yes
---
example:
X is Normally Distributed with a known mean of 55 and a standard deviation of 7 (variance 7^2) 
In simple words, 2 parameters define the normal (Gaussian) distribution: the mean and the variance (sd ^ 2)

```{r}
# draw 1 million observations from the above mentioned normal distribution
randsamp = rnorm(n = 1000000, mean = 55, sd = 7)
hist(randsamp, breaks = 500)


# probability than X is lower than 80
abline(v = 80, col = 2, lty = 2, lwd = 3)
pnorm(q = 80,mean = 55,sd = 5,lower.tail = T) #less than 80


# probability than X is lower than 55
abline(v = 55, col = "white", lty = 2, lwd = 3)
pnorm(q = 55, mean = 55, sd = 5, lower.tail = T)

#area above a q = 55 is 50% of total area..

# probability than X is higher than 56
pnorm(q=56,mean=55,sd=5,lower.tail=F)



#standard normal Z
#mean = 0, sd=1
#P than Z is lower than 1
randsamp = rnorm(n = 1000000, mean = 0, sd = 1)
hist(randsamp, breaks = 500)
abline(v = 1, col = "green", lty = 2, lwd = 3)
pnorm(q = 1, mean = 0, sd = 1, lower.tail = F)
#16% is aprroximately the area above 1

```



x is Normally Distributed with a known mean of 75 and a standard deviation of 5 (variance 5^2 = 25)
```{r}
#let's create a sequence of values that should cover large part of the normal distribution
x = seq(from = 55,to = 95,by = 0.25)
x

#let's compute the value of the probability density functions for these X values
dens = dnorm(x, mean = 75, sd = 5)
dens

plot(x, dens)

plot(x, dens, type = "l")

plot(x, dens, type = "l", main = "X~Normal: Mean = 75, SD = 5",xlab = "x", ylab = "Propability Density", las = 1)

#add vertical line
abline(v = 75, col = 2, lty = 2, lwd = 3)


```


-----> IN-CLASS EXERCISE



+++++++++++++++++
Edited by Simone Ciuti, University of Freiburg, 16/10/2015; 
Intended for the only purpose of teaching @ Freiburg University
+++++++++++++++++++++++++++++++++++++++++++++++++